---
title: "3_DGE"
author: "Zeqi Niu"
date: "6/29/2021"
output:
  html_document:
    toc: true
    toc_float: true
    theme: readable
    highlight: tango
    code_folding: show
---

*Goal*:
This script divides the dataset into different groups and do DGE analysis. The code will reset the dataset before every comparison range change. 

# Load in data and filtering
Takes normalized exprSet and cleaned up sdata
```{r}
library(limma)
library(tidyverse)
# (optional) if script1 "1_data_import_and_clean_up" was run before and you have the csv files, you can start this with:
exprSet_raw <-  read.csv("./exprSet_step1.csv", check.names = FALSE, row.names = 1)
sdata <- read.csv("./sdata_durva.csv", check.names = FALSE, row.names = 1)
```

# Filtering out the low express gene  
Because of the low dynamic range of microarray data, a lot of the low epxressing genes will be ranked low, and their DEG does not really have actual meanings, they are only low expressing genes. 

```{r}
medianFilter <- TRUE
sdFilter <- FALSE
sdFilterCutoff <- 0.05

# medianFilter: half of the median of the median list of all genes. 
if(medianFilter) {
  gene.median <- apply(exprSet_raw,1,median)
  halfMed <- median(gene.median)
  ind <- which(gene.median>halfMed)
  exprSet <- exprSet_raw[ind, ]
}

#sdFilter
if(sdFilter){
  gene.mean1 <- apply(exprSet_raw,1,mean)
  gene.sd1 <- apply(exprSet_raw,1,sd)
  gene.cv1 <- gene.sd1/gene.mean1
  ind <- which(gene.cv1> sdFilterCutoff  )
  exprSet <- exprSet_raw[ind, ]
}
```

for all these analysis for the first paper, we are going to focus only on Cells not Exosome samples for now. 

```{r}
sdata <- sdata[sdata$`Sample Type` == "Cell",]
exprSet_backup <- exprSet
sdata_backup <- sdata
```

# categorical variables 

some nots on  the common data reset and cleanup paragraph:
adjust sdata and exprSet into the correct comparison, for example, do we want to compare in all samples?or just pull out the cancer patient samples?  
Note: when changing different comparison ranges, one need to come back here to change comprange and exprSet, sdata in order for everything to match up.  
Set variable comprange to be indicate what is the comparison happening within: 
"whole": entire dataset including healthy and cancer;   
"cancer", cancer only;   
"bwk4": only baseline and wk4 patients;   
"ba": only baseline and after patients (may contain the wk4 as well)  
"base": baseline patient only  
"mid": wk4 patient only  
"after": 1month patient only  
"progression": all progression patients  
"stable": all stable patients  


##  cancer vs healthy vs cell line  
comprange: whole; compare group: cell line vs healthy, healthy vs tumor, cell line vs tumor. 


```{r}
comprange <- "whole"
exprSet <- exprSet_backup
sdata <- sdata_backup
sdata <- sdata
exprSet <- exprSet[,colnames(exprSet) %in% sdata$`Sample Name`]
dim(sdata)
dim(exprSet)
length(unique(sdata$PatientID))
```
```{r}
if (comprange == "whole"){
  groups = sdata$Type
  table(groups) 
  f = factor(groups,levels=c("Tumor","Healthy","Cell line")) # change the levels based on the table results from last line
  design = model.matrix(~ 0 + f)
  colnames(design)=c("Tumor","Healthy","Cellline") #(optional) would recommend naming it same as the levels. 
  data.fit = lmFit(exprSet, design)
  contrast.matrix = makeContrasts(Cellline - Healthy, Healthy-Tumor, Cellline-Tumor, levels = design)
  data.fit.con = contrasts.fit(data.fit,contrast.matrix)
  data.fit.eb = eBayes(data.fit.con)
  
  # results
  result <- decideTests(data.fit.eb)
  print(summary(result))
  top_genes <- topTable(data.fit.eb, coef = 2, number = 100, adjust = "fdr")
}
```
##  Time points  
comprange: cancer; compare group: baseline vs wk4, baseline vs after, wk4 vs after
```{r}
#reset
comprange <- "cancer" 
exprSet <- exprSet_backup
sdata <- sdata_backup
# new comprange
sdata <- sdata[sdata$Type == "Tumor",]
exprSet <- exprSet[,colnames(exprSet) %in% sdata$`Sample Name`]
#double check dimension
dim(sdata)
dim(exprSet)
length(unique(sdata$PatientID))
```
```{r}
if (comprange == "cancer"){
  groups = sdata$TimePoint
  table(groups) 
  f = factor(groups,levels=c("After","Base","Mid")) # change the levels based on the table results from last line
  design = model.matrix(~ 0 + f)
  colnames(design)=c("After","Base","Mid") #(optional) would recommend naming it same as the levels. 
  data.fit = lmFit(exprSet, design)
  contrast.matrix = makeContrasts(Base - Mid, Base-After, Mid-After, levels = design)
  data.fit.con = contrasts.fit(data.fit,contrast.matrix)
  data.fit.eb = eBayes(data.fit.con)
  
  # results
  result <- decideTests(data.fit.eb)
  print(summary(result))
  top_genes <- topTable(data.fit.eb, coef = 1, number = Inf, adjust = "fdr")
}
```

##  Progression status  
comprange: cancer; compare group: progression vs stable
```{r}
if (comprange == "cancer"){
  groups = sdata$Status
  table(groups) 
  f = factor(groups,levels=c("progression","stable")) # change the levels based on the table results from last line
  design = model.matrix(~ 0 + f)
  colnames(design)=c("progression","stable") #(optional) would recommend naming it same as the levels. 
  data.fit = lmFit(exprSet, design)
  contrast.matrix = makeContrasts(progression - stable, levels = design)
  data.fit.con = contrasts.fit(data.fit,contrast.matrix)
  data.fit.eb = eBayes(data.fit.con)
  
  # results
  result <- decideTests(data.fit.eb)
  print(summary(result))
  top_genes <- topTable(data.fit.eb, coef = 1, number = Inf, adjust = "fdr")
}
```

##  CTC level  
comprange: cancer minus some samples(didn't have CTC number); 

```{r}
if (comprange == "cancer"){
  # subsetting to patients who had CTC count data
  sdata <- sdata[!is.na(sdata$`CTC level`),]
  exprSet <- exprSet[,colnames(exprSet) %in% sdata$`Sample Name`]
  dim(sdata)
  dim(exprSet)
  
  # DEG analysis
  groups = sdata$`CTC level`
  table(groups) 
  f = factor(groups,levels=c("high","low","mid")) # change the levels based on the table results from last line
  design = model.matrix(~ 0 + f)
  colnames(design)=c("high","low", "mid") #(optional) would recommend naming it same as the levels. 
  data.fit = lmFit(exprSet, design)
  contrast.matrix = makeContrasts(high - low, mid - low, mid- high, levels = design)
  data.fit.con = contrasts.fit(data.fit,contrast.matrix)
  data.fit.eb = eBayes(data.fit.con)
  
  # results
  result <- decideTests(data.fit.eb)
  print(summary(result))
  top_genes <- topTable(data.fit.eb, coef = 1, number = Inf, adjust = "fdr", p.value = 0.05)
  top_genes_lfc <-top_genes[order(abs(top_genes$logFC),decreasing = TRUE),]
  print(head(top_genes_lfc, 10))
}
```

*for generating Figure 5D*
```{r}
exprSet <- exprSet_backup
sdata <- sdata_backup
# new comprange
sdata <- sdata[sdata$Type == "Tumor",]
sdata <- sdata[order(sdata$`CTC level`),]
sdata <- sdata[!is.na(sdata$`CTC level`),]
# this is for Figure 5 D
sdata_high <- sdata[sdata$`CTC level` == "high",]
sdata_low <- sdata[sdata$`CTC level` == "low",]
exprSet_high <- exprSet[,colnames(exprSet) %in% sdata_high$`Sample Name`]
exprSet_low <- exprSet[,colnames(exprSet) %in% sdata_low$`Sample Name`]
G3BP1_high <- exprSet_high[rownames(exprSet_high) == "G3BP1",]; G3BP1_high$group <- "high"
G3BP1_low <- exprSet_low[rownames(exprSet_low) == "G3BP1",]; G3BP1_low$group <- "low"

write.csv(G3BP1_high,"G3BP1_CTChigh.csv")
write.csv(G3BP1_low,"G3BP1_CTClow.csv")
```

  
##  progression prediction at each single time point 

### base
comprange: base; compare group: progression vs stable
```{r}
#reset
comprange <- "base" 
exprSet <- exprSet_backup
sdata <- sdata_backup
# new comprange
sdata <- sdata[sdata$TimePoint=="Base",]        
exprSet <- exprSet[,colnames(exprSet) %in% sdata$`Sample Name`]
#double check dimension
dim(sdata)
dim(exprSet)
```
```{r}
if (comprange == "base"){
  groups = sdata$Status
  table(groups) 
  f = factor(groups,levels=c("progression","stable")) # change the levels based on the table results from last line
  design = model.matrix(~ 0 + f)
  colnames(design)=c("progression","stable") 
  data.fit = lmFit(exprSet, design)
  contrast.matrix = makeContrasts(progression - stable, levels = design)
  data.fit.con = contrasts.fit(data.fit,contrast.matrix)
  data.fit.eb = eBayes(data.fit.con)
  
  # results
  result <- decideTests(data.fit.eb, p.value = 0.05)
  print(summary(result))
}

```
### wk4
comprange: wk4; compare group: progression vs stable  
```{r}
#reset
comprange <- "mid" 
exprSet <- exprSet_backup
sdata <- sdata_backup
# new comprange
sdata <- sdata[sdata$TimePoint=="Mid",]        
exprSet <- exprSet[,colnames(exprSet) %in% sdata$`Sample Name`]
#double check dimension
dim(sdata)
dim(exprSet)
```
```{r}
if (comprange == "mid"){
  groups = sdata$Status
  table(groups) 
  f = factor(groups,levels=c("progression","stable")) # change the levels based on the table results from last line
  design = model.matrix(~ 0 + f)
  colnames(design)=c("progression","stable") 
  data.fit = lmFit(exprSet, design)
  contrast.matrix = makeContrasts(progression - stable, levels = design)
  data.fit.con = contrasts.fit(data.fit,contrast.matrix)
  data.fit.eb = eBayes(data.fit.con)
  
  # results
  result <- decideTests(data.fit.eb, p.value = 0.05)
  print(summary(result))
}

```
### after
comprange: after; compare group: progression vs stable  
```{r}
#reset
comprange <- "after" 
exprSet <- exprSet_backup
sdata <- sdata_backup
# new comprange
sdata <- sdata[sdata$TimePoint=="After",]        
exprSet <- exprSet[,colnames(exprSet) %in% sdata$`Sample Name`]
#double check dimension
dim(sdata)
dim(exprSet)
```
```{r}
if (comprange == "after"){
  groups = sdata$Status
  table(groups) 
  f = factor(groups,levels=c("progression","stable")) # change the levels based on the table results from last line
  design = model.matrix(~ 0 + f)
  colnames(design)=c("progression","stable") 
  data.fit = lmFit(exprSet, design)
  contrast.matrix = makeContrasts(progression - stable, levels = design)
  data.fit.con = contrasts.fit(data.fit,contrast.matrix)
  data.fit.eb = eBayes(data.fit.con)
  
  # results
  result <- decideTests(data.fit.eb, p.value = 0.05)
  print(summary(result))
}

```


## timepoint in all patient group

### paired-analysis 

comprange: bwk4; compare group: baseline vs. wk4  
```{r}
#reset
comprange <- "bwk4" 
exprSet <- exprSet_backup
sdata <- sdata_backup

# new comprange
sdata_base <- sdata[sdata$TimePoint=="Base",]
sdata_mid <- sdata[sdata$TimePoint=="Mid",]
sdata_after <- sdata[sdata$TimePoint=="After",]
patient_id <- intersect(sdata_base$PatientID, sdata_mid$PatientID)
sdata <- sdata[sdata$PatientID %in% patient_id,]
sdata <- sdata[!sdata$TimePoint == "After",]
exprSet <- exprSet[,colnames(exprSet) %in% sdata$`Sample Name`]

#double check dimension
dim(sdata)
dim(exprSet)
```
```{r}
if (comprange == "bwk4"){
  groups = sdata$TimePoint
  table(groups) 
  f = factor(groups,levels=c("Base","Mid")) # change the levels based on the table results from last line
  ID = factor(sdata$PatientID)
  design = model.matrix(~ ID + f)
  data.fit = lmFit(exprSet, design)
  data.fit.eb = eBayes(data.fit)
  
  # results
  result <- decideTests(data.fit.eb, p.value = 0.05)
  print(summary(result))
  top_genes <- topTable(data.fit.eb, coef = "fMid", number = 50, adjust = "fdr", p.value = 0.05)
  write.csv(top_genes, "bw4_fold_change_DEGs_sig_genes_top50.csv")
  #top_genes <- top_genes[top_genes$adj.P.Val < 0.05,]
  #top_genes_lfc <-top_genes[order(abs(top_genes$logFC),decreasing = TRUE),]
  
  #rite.csv(top_genes_lfc, "./toptables/bwk4_paired_sortlfc_91.csv")
}

```

comprange: ba; compare group: baseline vs. after
```{r}
#reset
comprange <- "ba" 
exprSet <- exprSet_backup
sdata <- sdata_backup

# new comprange
sdata_base <- sdata[sdata$TimePoint=="Base",]
sdata_mid <- sdata[sdata$TimePoint=="Mid",]
sdata_after <- sdata[sdata$TimePoint=="After",]
patient_id <- intersect(sdata_base$PatientID, sdata_after$PatientID)
sdata <- sdata[sdata$PatientID %in% patient_id,]
sdata <- sdata[!sdata$TimePoint == "Mid",]
exprSet <- exprSet[,colnames(exprSet) %in% sdata$`Sample Name`]

#double check dimension
dim(sdata)
dim(exprSet)
```
```{r}
if (comprange == "ba"){
  groups = sdata$TimePoint
  table(groups) 
  f = factor(groups,levels=c("Base","After")) # change the levels based on the table results from last line
  ID = factor(sdata$PatientID)
  design = model.matrix(~ ID + f)
  data.fit = lmFit(exprSet, design)
  data.fit.eb = eBayes(data.fit)
  
  # results
  result <- decideTests(data.fit.eb, p.value = 0.05)
  print(summary(result))
  top_genes <- topTable(data.fit.eb, coef = "fAfter", number = 50, adjust = "fdr")
  top_genes_lfc <-top_genes[order(abs(top_genes$logFC),decreasing = TRUE),]
}

```

## timepoint in status subsets 
### progression patients

--------

*baseline vs mid*
comprange: progression (bm); compare group: baseline vs. mid;   

```{r}
# reset
comprange <- "progression" 
exprSet <- exprSet_backup
sdata <- sdata_backup
# new comprange
sdata <- sdata[sdata$Type == "Tumor",]
sdata <- sdata[sdata$Status=="progression",]
sdata <- sdata[!sdata$TimePoint == "After",]
exprSet <- exprSet[,colnames(exprSet) %in% sdata$`Sample Name`]

#double check dimension
dim(sdata)
dim(exprSet)
length(unique(sdata$PatientID))
```
#### unpaired analysis
```{r}
if (comprange == "progression"){
  groups = sdata$TimePoint
  table(groups) 
  f = factor(groups,levels=c("Base","Mid")) # change the levels based on the table results from last line
  ID = factor(sdata$PatientID)
  design = model.matrix(~ 1 + f)
  data.fit = lmFit(exprSet, design)
  data.fit.eb = eBayes(data.fit)
  
  # results
  result <- decideTests(data.fit.eb, p.value = 0.05)
  print(summary(result))
  top_genes <- topTable(data.fit.eb, coef = "fMid", number = Inf, adjust = "fdr")
  top_genes_lfc <-top_genes[order(abs(top_genes$logFC),decreasing = TRUE),]
  print(head(top_genes_lfc, 10))
}

```
```{r}
# This is for plotting in figure 5 
stable_gene_list <- read.csv("stable patients bw4 DEG signatures.csv")
stable_gene_list <- stable_gene_list$X
table(rownames(top_genes) %in% stable_gene_list)
stable_top_genes <- top_genes[rownames(top_genes) %in% stable_gene_list,]
write.csv(stable_top_genes, "progressing patients bw4 DEG stable signatures.csv")
```

#### paired analysis
```{r}
# get rid of samples with single time points
idcount <- sdata %>% count(PatientID)
idcount <- idcount[idcount$n > 1,]
sdata <- sdata[sdata$PatientID %in% idcount$PatientID,]
exprSet <- exprSet[,colnames(exprSet) %in% sdata$`Sample Name`]

#double check dimension
dim(sdata)
dim(exprSet)
```
```{r}
if (comprange == "progression"){
  groups = sdata$TimePoint
  table(groups) 
  f = factor(groups,levels=c("Base","Mid")) # change the levels based on the table results from last line
  ID = factor(sdata$PatientID)
  design = model.matrix(~ ID + f)
  data.fit = lmFit(exprSet, design)
  data.fit.eb = eBayes(data.fit)
  
  # results
  result <- decideTests(data.fit.eb, p.value = 0.05)
  print(summary(result))
  top_genes <- topTable(data.fit.eb, coef = "fMid", number = 50, adjust = "fdr")
  top_genes_lfc <-top_genes[order(abs(top_genes$logFC),decreasing = TRUE),]
  print(head(top_genes_lfc, 10))
}

```


----------

*baseline vs after*
comprange: progression (ba); compare group: baseline vs. after;   
```{r}
# reset
comprange <- "progression" 
exprSet <- exprSet_backup
sdata <- sdata_backup
# new comprange
sdata <- sdata[sdata$Type == "Tumor",]
sdata <- sdata[sdata$Status=="progression",]
sdata <- sdata[!sdata$TimePoint == "Mid",]
exprSet <- exprSet[,colnames(exprSet) %in% sdata$`Sample Name`]

#double check dimension
dim(sdata)
dim(exprSet)
length(unique(sdata$PatientID))
```
#### unpaired analysis
```{r}
if (comprange == "progression"){
  groups = sdata$TimePoint
  table(groups) 
  f = factor(groups,levels=c("Base","After")) # change the levels based on the table results from last line
  ID = factor(sdata$PatientID)
  design = model.matrix(~ 1 + f)
  data.fit = lmFit(exprSet, design)
  data.fit.eb = eBayes(data.fit)
  
  # results
  result <- decideTests(data.fit.eb, p.value = 0.05)
  print(summary(result))
  top_genes <- topTable(data.fit.eb, coef = "fAfter", number = 50, adjust = "fdr")
  top_genes_lfc <-top_genes[order(abs(top_genes$logFC),decreasing = TRUE),]
  print(head(top_genes_lfc, 10))
}

```
#### paired analysis
```{r}
# get rid of samples with single time points
idcount <- sdata %>% count(PatientID)
idcount <- idcount[idcount$n > 1,]
sdata <- sdata[sdata$PatientID %in% idcount$PatientID,]
exprSet <- exprSet[,colnames(exprSet) %in% sdata$`Sample Name`]

#double check dimension
dim(sdata)
dim(exprSet)
length(unique(sdata$PatientID))
```
```{r}
if (comprange == "progression"){
  groups = sdata$TimePoint
  table(groups) 
  f = factor(groups,levels=c("Base","After")) # change the levels based on the table results from last line
  ID = factor(sdata$PatientID)
  design = model.matrix(~ ID + f)
  data.fit = lmFit(exprSet, design)
  data.fit.eb = eBayes(data.fit)
  
  # results
  result <- decideTests(data.fit.eb, p.value = 0.05)
  print(summary(result))
  top_genes <- topTable(data.fit.eb, coef = "fAfter", number = 50, adjust = "fdr")
  top_genes_lfc <-top_genes[order(abs(top_genes$logFC),decreasing = TRUE),]
  print(head(top_genes_lfc, 10))
}

```

### stable patients

----------

*baseline vs mid*
comprange: stable (bm); compare group: baseline vs. mid  
```{r}
# reset
comprange <- "stable" 
exprSet <- exprSet_backup
sdata <- sdata_backup
# new comprange
sdata <- sdata[sdata$Type == "Tumor",]
sdata <- sdata[sdata$Status=="stable",]
sdata <- sdata[!sdata$TimePoint == "After",]
exprSet <- exprSet[,colnames(exprSet) %in% sdata$`Sample Name`]

#double check dimension
dim(sdata)
dim(exprSet)
length(unique(sdata$PatientID))
```
#### unpaired analysis
```{r}
if (comprange == "stable"){
  groups = sdata$TimePoint
  table(groups) 
  f = factor(groups,levels=c("Base","Mid")) # change the levels based on the table results from last line
  #ID = factor(sdata$PatientID)
  design = model.matrix( ~1 + f)
  data.fit = lmFit(exprSet, design)
  data.fit.eb = eBayes(data.fit)
  
  # results
  result <- decideTests(data.fit.eb, p.value = 0.05)
  print(summary(result))
  top_genes <- topTable(data.fit.eb, coef = "fMid", number = 50, adjust = "fdr")
  top_genes <- top_genes[top_genes$adj.P.Val < 0.05,]
  top_genes_lfc <-top_genes[order(abs(top_genes$logFC),decreasing = TRUE),]
  print(head(top_genes_lfc, 10))
  write.csv(top_genes_lfc, "stable patients bw4 DEG signatures.csv")
}

```
#### paired analysis
comprange: stable; compare group: baseline vs. mid  
```{r}
# get rid of samples with single time points
idcount <- sdata %>% count(PatientID)
idcount <- idcount[idcount$n > 1,]
sdata <- sdata[sdata$PatientID %in% idcount$PatientID,]
exprSet <- exprSet[,colnames(exprSet) %in% sdata$`Sample Name`]

#double check dimension
dim(sdata)
dim(exprSet)
length(unique(sdata$PatientID))
```
```{r}
if (comprange == "stable"){
  groups = sdata$TimePoint
  table(groups) 
  f = factor(groups,levels=c("Base","Mid")) # change the levels based on the table results from last line
  ID = factor(sdata$PatientID)
  design = model.matrix( ~ID + f)
  data.fit = lmFit(exprSet, design)
  data.fit.eb = eBayes(data.fit)
  
  # results
  result <- decideTests(data.fit.eb, p.value = 0.05)
  print(summary(result))
  top_genes <- topTable(data.fit.eb, coef = "fMid", number = 50, adjust = "fdr")
  top_genes_lfc <-top_genes[order(abs(top_genes$logFC),decreasing = TRUE),]
  print(head(top_genes_lfc, 10))
}
```

----------

*baseline vs after*


comprange: stable (ba); compare group: baseline vs. after   
```{r}
# reset
comprange <- "stable" 
exprSet <- exprSet_backup
sdata <- sdata_backup
# new comprange
sdata <- sdata[sdata$Type == "Tumor",]
sdata <- sdata[sdata$Status=="stable",]
sdata <- sdata[!sdata$TimePoint == "Mid",]
exprSet <- exprSet[,colnames(exprSet) %in% sdata$`Sample Name`]

#double check dimension
dim(sdata)
dim(exprSet)
length(unique(sdata$PatientID))
```
#### unpaired analysis
```{r}
if (comprange == "stable"){
  groups = sdata$TimePoint
  table(groups) 
  f = factor(groups,levels=c("Base","After")) # change the levels based on the table results from last line
  #ID = factor(sdata$PatientID)
  design = model.matrix( ~1 + f)
  data.fit = lmFit(exprSet, design)
  data.fit.eb = eBayes(data.fit)
  
  # results
  result <- decideTests(data.fit.eb, p.value = 0.05)
  print(summary(result))
  top_genes <- topTable(data.fit.eb, coef = "fAfter", number = 50, adjust = "fdr")
  top_genes_lfc <-top_genes[order(abs(top_genes$logFC),decreasing = TRUE),]
  print(head(top_genes_lfc, 10))
}

```
#### paired analysis
comprange: stable(ba); compare group: baseline vs. after  
```{r}
# get rid of samples with single time points
idcount <- sdata %>% count(PatientID)
idcount <- idcount[idcount$n > 1,]
sdata <- sdata[sdata$PatientID %in% idcount$PatientID,]
exprSet <- exprSet[,colnames(exprSet) %in% sdata$`Sample Name`]

#double check dimension
dim(sdata)
dim(exprSet)
length(unique(sdata$PatientID))
```
```{r}
if (comprange == "stable"){
  groups = sdata$TimePoint
  table(groups) 
  f = factor(groups,levels=c("Base","After")) # change the levels based on the table results from last line
  ID = factor(sdata$PatientID)
  design = model.matrix( ~ID + f)
  data.fit = lmFit(exprSet, design)
  data.fit.eb = eBayes(data.fit)
  
  # results
  result <- decideTests(data.fit.eb, p.value = 0.05)
  print(summary(result))
  top_genes <- topTable(data.fit.eb, coef = "fAfter", number = 50, adjust = "fdr")
  top_genes_lfc <-top_genes[order(abs(top_genes$logFC),decreasing = TRUE),]
  print(head(top_genes_lfc, 10))
}
```

# PDL1 high vs low comparison
## all patients
```{r}
comprange <- "whole"
sdata <- sdata_backup
exprSet <- exprSet_raw[,colnames(exprSet_raw) %in% sdata$`Sample Name`]
dim(sdata)
dim(exprSet)
length(unique(sdata$PatientID))
#median filter gets rid of PDL1
midExpr <-  as.numeric(quantile(exprSet[rownames(exprSet) == "CD274",], probs = 0.5))
PDL1 <- t(as.data.frame(exprSet[rownames(exprSet) == "CD274",]))
sdata <- merge(sdata, PDL1, by.x = "Sample Name", by.y = 0)
sdata$CD274 <- ifelse(sdata$CD274 >= midExpr, 'High','Low')
```
```{r}
if (comprange == "whole"){
  groups = sdata$CD274
  table(groups) 
  f = factor(groups,levels=c("High","Low")) # change the levels based on the table results from last line
  design = model.matrix(~ 0 + f)
  colnames(design)=c("High","Low") #(optional) would recommend naming it same as the levels. 
  data.fit = lmFit(exprSet, design)
  contrast.matrix = makeContrasts(High - Low, levels = design)
  data.fit.con = contrasts.fit(data.fit,contrast.matrix)
  data.fit.eb = eBayes(data.fit.con)
  
  # results
  result <- decideTests(data.fit.eb)
  print(summary(result))
  top_genes <- topTable(data.fit.eb, number = 100, adjust = "fdr")
}
```

## base patients
```{r}
#reset
comprange <- "base" 
sdata <- sdata_backup
sdata <- sdata[sdata$TimePoint=="Base",]
exprSet <- exprSet_raw[,colnames(exprSet_raw) %in% sdata$`Sample Name`]
dim(sdata)
dim(exprSet)
length(unique(sdata$PatientID))
#median filter gets rid of PDL1
midExpr <-  as.numeric(quantile(exprSet[rownames(exprSet) == "CD274",], probs = 0.5))
PDL1 <- t(as.data.frame(exprSet[rownames(exprSet) == "CD274",]))
sdata <- merge(sdata, PDL1, by.x = "Sample Name", by.y = 0)
sdata$CD274 <- ifelse(sdata$CD274 >= midExpr, 'High','Low')
```
```{r}
if (comprange == "base"){
  groups = sdata$CD274
  table(groups) 
  f = factor(groups,levels=c("High","Low")) # change the levels based on the table results from last line
  design = model.matrix(~ 0 + f)
  colnames(design)=c("High","Low") #(optional) would recommend naming it same as the levels. 
  data.fit = lmFit(exprSet, design)
  contrast.matrix = makeContrasts(High - Low, levels = design)
  data.fit.con = contrasts.fit(data.fit,contrast.matrix)
  data.fit.eb = eBayes(data.fit.con)
  
  # results
  result <- decideTests(data.fit.eb)
  print(summary(result))
  top_genes <- topTable(data.fit.eb, number = 100, adjust = "fdr")
}
```

## wk4 patients
```{r}
#reset
comprange <- "mid" 
sdata <- sdata_backup
sdata <- sdata[sdata$TimePoint=="Mid",]
exprSet <- exprSet_raw[,colnames(exprSet_raw) %in% sdata$`Sample Name`]
dim(sdata)
dim(exprSet)
length(unique(sdata$PatientID))
#median filter gets rid of PDL1
midExpr <-  as.numeric(quantile(exprSet[rownames(exprSet) == "CD274",], probs = 0.5))
PDL1 <- t(as.data.frame(exprSet[rownames(exprSet) == "CD274",]))
sdata <- merge(sdata, PDL1, by.x = "Sample Name", by.y = 0)
sdata$CD274 <- ifelse(sdata$CD274 >= midExpr, 'High','Low')
```
```{r}
if (comprange == "mid"){
  groups = sdata$CD274
  table(groups) 
  f = factor(groups,levels=c("High","Low")) # change the levels based on the table results from last line
  design = model.matrix(~ 0 + f)
  colnames(design)=c("High","Low") #(optional) would recommend naming it same as the levels. 
  data.fit = lmFit(exprSet, design)
  contrast.matrix = makeContrasts(High - Low, levels = design)
  data.fit.con = contrasts.fit(data.fit,contrast.matrix)
  data.fit.eb = eBayes(data.fit.con)
  
  # results
  result <- decideTests(data.fit.eb)
  print(summary(result))
  top_genes <- topTable(data.fit.eb, number = 100, adjust = "fdr")
}
```

## after patients
```{r}
#reset
comprange <- "after" 
sdata <- sdata_backup
sdata <- sdata[sdata$TimePoint=="After",]
exprSet <- exprSet_raw[,colnames(exprSet_raw) %in% sdata$`Sample Name`]
dim(sdata)
dim(exprSet)
length(unique(sdata$PatientID))
#median filter gets rid of PDL1
midExpr <-  as.numeric(quantile(exprSet[rownames(exprSet) == "CD274",], probs = 0.5))
PDL1 <- t(as.data.frame(exprSet[rownames(exprSet) == "CD274",]))
sdata <- merge(sdata, PDL1, by.x = "Sample Name", by.y = 0)
sdata$CD274 <- ifelse(sdata$CD274 >= midExpr, 'High','Low')
```
```{r}
if (comprange == "after"){
  groups = sdata$CD274
  table(groups) 
  f = factor(groups,levels=c("High","Low")) # change the levels based on the table results from last line
  design = model.matrix(~ 0 + f)
  colnames(design)=c("High","Low") #(optional) would recommend naming it same as the levels. 
  data.fit = lmFit(exprSet, design)
  contrast.matrix = makeContrasts(High - Low, levels = design)
  data.fit.con = contrasts.fit(data.fit,contrast.matrix)
  data.fit.eb = eBayes(data.fit.con)
  
  # results
  result <- decideTests(data.fit.eb)
  print(summary(result))
  top_genes <- topTable(data.fit.eb, number = 100, adjust = "fdr")
}
```