---
title: "3_DGE"
author: "Zeqi Niu"
date: "6/29/2021"
output:
  html_document:
    toc: true
    toc_float: true
    theme: readable
    highlight: tango
    code_folding: show
---

*Goal*:
This script divides the dataset into different groups and do DGE analysis. The code will reset the dataset before every comparison range change. 

# Load in data and filtering
Takes normalized exprSet and cleaned up sdata
```{r}
library(limma)
library(tidyverse)
# (optional) if script1 "1_data_import_and_clean_up" was run before and you have the csv files, you can start this with:
exprSet_raw <-  read.csv("./exprSet_step1.csv", check.names = FALSE, row.names = 1)
sdata <- read.csv("./sdata_durva.csv", check.names = FALSE, row.names = 1)
```

# Filtering out the low express gene  
Because of the low dynamic range of microarray data, a lot of the low epxressing genes will be ranked low, and their DEG does not really have actual meanings, they are only low expressing genes. 

```{r}
medianFilter <- TRUE
sdFilter <- FALSE
sdFilterCutoff <- 0.05

# medianFilter: half of the median of the median list of all genes. 
if(medianFilter) {
  gene.median <- apply(exprSet_raw,1,median)
  halfMed <- median(gene.median)
  ind <- which(gene.median>halfMed)
  exprSet <- exprSet_raw[ind, ]
}

#sdFilter
if(sdFilter){
  gene.mean1 <- apply(exprSet_raw,1,mean)
  gene.sd1 <- apply(exprSet_raw,1,sd)
  gene.cv1 <- gene.sd1/gene.mean1
  ind <- which(gene.cv1> sdFilterCutoff  )
  exprSet <- exprSet_raw[ind, ]
}
```

for all these analysis for the first paper, we are going to focus only on Cells not Exosome samples for now. 

```{r}
sdata <- sdata[sdata$`Sample Type` == "Cell",]
exprSet_backup <- exprSet
sdata_backup <- sdata
```

# categorical variables 

some nots on  the common data reset and cleanup paragraph:
adjust sdata and exprSet into the correct comparison, for example, do we want to compare in all samples?or just pull out the cancer patient samples?  
Note: when changing different comparison ranges, one need to come back here to change comprange and exprSet, sdata in order for everything to match up.  
Set variable comprange to be indicate what is the comparison happening within: 
"whole": entire dataset including healthy and cancer;   
"cancer", cancer only;   
"bwk4": only baseline and wk4 patients;   
"ba": only baseline and after patients (may contain the wk4 as well)  
"base": baseline patient only  
"mid": wk4 patient only  
"after": 1month patient only  
"progression": all progression patients  
"stable": all stable patients  

## Figure 3 patients only 
```{r}
comprange <-  "figure3"
sdata <- sdata[sdata$Figure3_binary == 1,]
sdata <- sdata[!is.na(sdata$Well),]
print("patients included:"); print(unique(sdata1$PatientID))
exprSet <- exprSet[,colnames(exprSet) %in% sdata$`Sample Name`]
dim(sdata)
dim(exprSet)
length(unique(sdata$PatientID))
```

##  Progression status  
comprange: cancer; compare group: progression vs stable
```{r}
if (comprange == "figure3"){
  groups = sdata$Status
  table(groups) 
  f = factor(groups,levels=c("progression","stable")) # change the levels based on the table results from last line
  design = model.matrix(~ 0 + f)
  colnames(design)=c("progression","stable") #(optional) would recommend naming it same as the levels. 
  data.fit = lmFit(exprSet, design)
  contrast.matrix = makeContrasts(progression - stable, levels = design)
  data.fit.con = contrasts.fit(data.fit,contrast.matrix)
  data.fit.eb = eBayes(data.fit.con)
  
  # results
  result <- decideTests(data.fit.eb)
  print(summary(result))
  top_genes <- topTable(data.fit.eb, coef = 1, number = Inf, adjust = "fdr")
}
```
