---
title: "S4_survival_gene"
author: "Zeqi Niu"
date: "10/10/2021"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r}
library(survival)
library(readxl)
library(tidyverse)
library(survival)
library(RegParallel)
```
# load in pre calculated PFS data
```{r}
exprSet <-  read.csv("./exprSet_step1.csv", check.names = FALSE, row.names = 1)
exprSet_backup <- exprSet
sdata <- read.csv("./sdata.csv", check.names = FALSE, row.names = 1)
pfs <- read_excel("../matching_PFStable_Dec_newversion.xlsx")
```

# data clean up
```{r}
pfs1 <- separate(pfs, `patient ID`, c("year", "um", "id"), sep = " " )
pfs1$id <- as.factor(as.numeric(pfs1$id))
sdata1 <- sdata[,-14:-37]
sdata1$PurificationTechnology <- NULL
sdata1$`Sample Source` <- NULL
sdata1$`Plate Name` <- NULL

pfs2 <- pfs1[ , c("id", "PFS Days","PFS month","right censor","progression event")]
```
for checking S1
```{r}
pdata <- merge(sdata1, pfs2, by.x = "PatientID", by.y = "id")
```


# start with simple one timepoint gene expression PFS prediction. 


## only looking at baseline data

```{r}
sdata1_base <- sdata1[sdata1$TimePoint == "Base",]
pdata_base <- merge(sdata1_base, pfs2, by.x = "PatientID", by.y = "id")
```
```{r}
exprSet <- exprSet_backup
exprSet_base <- exprSet[, colnames(exprSet) %in% pdata_base$`Sample Name` ]
exprSet_base1 <- t(exprSet_base)
coxdata <- merge(pdata_base, exprSet_base1, by.x ="Sample Name", by.y = 0)

colnames <- make.names(colnames(coxdata))
colnames(coxdata) <- colnames
```

# PDL1 high vs low
```{r}
gene <- "CD274"
survplotdata <- coxdata[,c('PFS.month', 'progression.event',paste(gene))]

# set Z-scale cut-offs for high and low expression
highExpr <- as.numeric(quantile(survplotdata[,paste(gene)], probs = 0.75))
lowExpr <- as.numeric(quantile(survplotdata[,paste(gene)], probs = 0.25))
midExpr <-  as.numeric(quantile(survplotdata[,paste(gene)], probs = 0.5))

survplotdata[,paste(gene)] <- ifelse(survplotdata[,paste(gene)] >= midExpr, 'High','Low')


# relevel the factors to have mid as the ref level
survplotdata[,paste(gene)] <- factor(survplotdata[,paste(gene)],
                              levels = c('Low', 'High'))

library(survminer)


ggsurvplot(survfit(Surv(PFS.month, progression.event) ~ CD274,
                   data = survplotdata),
            data = survplotdata,
            pval = TRUE,
            break.time.by = 5,
            ggtheme = theme_minimal(),
            risk.table.y.text.col = TRUE,
            risk.table.y.text = FALSE)
```

```{r}
res <- RegParallel(
  data = coxdata,
  formula = 'Surv(`PFS.month`, `progression.event`) ~ [*]',
  FUN = function(formula, data)
    coxph(formula = formula,
          data = data,
          ties = 'breslow',
          singular.ok = TRUE),
  FUNtype = 'coxph',
  variables = colnames(coxdata)[15:ncol(coxdata)],
  blocksize = 2000,
  cores = 2,
  nestedParallel = FALSE,
  conflevel = 95,
  p.adjust = 'BH')
res <- res[!is.na(res$P),]
res

res <- res[order(res$LogRank, decreasing = FALSE),]
final_base <- subset(res, LogRank < 0.01)
write.csv(final_base, "baseline expression with survival.csv")
```

```{r}
gene <- "ZNF205"
survplotdata <- coxdata[,c('PFS.month', 'progression.event',paste(gene))]

# set Z-scale cut-offs for high and low expression
highExpr <- as.numeric(quantile(survplotdata[,paste(gene)], probs = 0.75))
lowExpr <- as.numeric(quantile(survplotdata[,paste(gene)], probs = 0.25))

survplotdata[,paste(gene)] <- ifelse(survplotdata[,paste(gene)] >= highExpr, 'High',
                              ifelse(survplotdata[,paste(gene)] <= lowExpr, 'Low', 'Mid'))


# relevel the factors to have mid as the ref level
survplotdata[,paste(gene)] <- factor(survplotdata[,paste(gene)],
                              levels = c('Mid', 'Low', 'High'))

library(survminer)


ggsurvplot(survfit(Surv(PFS.month, progression.event) ~ ZNF205,
                   data = survplotdata),
            data = survplotdata,
            pval = TRUE,
            break.time.by = 5,
            ggtheme = theme_minimal(),
            risk.table.y.text.col = TRUE,
            risk.table.y.text = FALSE)
```

```{r}
top_DEGs_91 <- read.csv("./toptables/bwk4_paired_sortlfc_91.csv")
top_DEGs_7 <- read.csv("stable patients bw4 DEG signatures.csv")
intersect(final_base$Variable, top_DEGs_91$X)
``` 

## only looking at wk4 data

```{r}
sdata1_wk4 <- sdata1[sdata1$TimePoint == "Mid",]
pdata_wk4 <- merge(sdata1_wk4, pfs2, by.x = "PatientID", by.y = "id")
```
```{r}
exprSet <- exprSet_backup
exprSet_wk4 <- exprSet[, colnames(exprSet) %in% pdata_wk4$`Sample Name` ]
exprSet_wk41 <- t(exprSet_wk4)
coxdata <- merge(pdata_wk4, exprSet_wk41, by.x ="Sample Name", by.y = 0)

colnames <- make.names(colnames(coxdata))
colnames(coxdata) <- colnames
coxdata[1:10,1:20]
```
# PDL1 high vs low
```{r}
gene <- "CD274"
survplotdata <- coxdata[,c('PFS.month', 'progression.event',paste(gene))]

# set Z-scale cut-offs for high and low expression
highExpr <- as.numeric(quantile(survplotdata[,paste(gene)], probs = 0.75))
lowExpr <- as.numeric(quantile(survplotdata[,paste(gene)], probs = 0.25))
midExpr <-  as.numeric(quantile(survplotdata[,paste(gene)], probs = 0.5))

survplotdata[,paste(gene)] <- ifelse(survplotdata[,paste(gene)] >= midExpr, 'High','Low')


# relevel the factors to have mid as the ref level
survplotdata[,paste(gene)] <- factor(survplotdata[,paste(gene)],
                              levels = c('Low', 'High'))

library(survminer)


ggsurvplot(survfit(Surv(PFS.month, progression.event) ~ CD274,
                   data = survplotdata),
            data = survplotdata,
            pval = TRUE,
            break.time.by = 5,
            ggtheme = theme_minimal(),
            risk.table.y.text.col = TRUE,
            risk.table.y.text = FALSE)
```
```{r}
res <- RegParallel(
  data = coxdata,
  formula = 'Surv(`PFS.month`, `progression.event`) ~ [*]',
  FUN = function(formula, data)
    coxph(formula = formula,
          data = data,
          ties = 'breslow',
          singular.ok = TRUE),
  FUNtype = 'coxph',
  variables = colnames(coxdata)[15:ncol(coxdata)],
  blocksize = 2000,
  cores = 2,
  nestedParallel = FALSE,
  conflevel = 95,
  p.adjust = 'BH')
res <- res[!is.na(res$P),]
res

res <- res[order(res$LogRank, decreasing = FALSE),]
final_wk4 <- subset(res, LogRank < 0.01)
intersect(final_wk4$Variable, final_base$Variable)
length(setdiff(final_wk4$Variable, final_base$Variable))
write.csv(final_wk4, "wk4 expression with survival.csv")
```

## only looking at after data

```{r}
sdata1_after <- sdata1[sdata1$TimePoint == "After",]
pdata_after <- merge(sdata1_after, pfs2, by.x = "PatientID", by.y = "id")
```
```{r}
exprSet <- exprSet_backup
exprSet_after <- exprSet[, colnames(exprSet) %in% pdata_after$`Sample Name` ]
exprSet_after1 <- t(exprSet_after)
coxdata <- merge(pdata_after, exprSet_after1, by.x ="Sample Name", by.y = 0)

colnames <- make.names(colnames(coxdata))
colnames(coxdata) <- colnames
coxdata[1:10,1:20]
```
# PDL1 high vs low
```{r}
gene <- "CD274"
survplotdata <- coxdata[,c('PFS.month', 'progression.event',paste(gene))]

# set Z-scale cut-offs for high and low expression
highExpr <- as.numeric(quantile(survplotdata[,paste(gene)], probs = 0.75))
lowExpr <- as.numeric(quantile(survplotdata[,paste(gene)], probs = 0.25))
midExpr <-  as.numeric(quantile(survplotdata[,paste(gene)], probs = 0.5))

survplotdata[,paste(gene)] <- ifelse(survplotdata[,paste(gene)] >= midExpr, 'High','Low')


# relevel the factors to have mid as the ref level
survplotdata[,paste(gene)] <- factor(survplotdata[,paste(gene)],
                              levels = c('Low', 'High'))

library(survminer)


ggsurvplot(survfit(Surv(PFS.month, progression.event) ~ CD274,
                   data = survplotdata),
            data = survplotdata,
            pval = TRUE,
            break.time.by = 5,
            ggtheme = theme_minimal(),
            risk.table.y.text.col = TRUE,
            risk.table.y.text = FALSE)
```
```{r}
res <- RegParallel(
  data = coxdata,
  formula = 'Surv(`PFS.month`, `progression.event`) ~ [*]',
  FUN = function(formula, data)
    coxph(formula = formula,
          data = data,
          ties = 'breslow',
          singular.ok = TRUE),
  FUNtype = 'coxph',
  variables = colnames(coxdata)[15:ncol(coxdata)],
  blocksize = 2000,
  cores = 2,
  nestedParallel = FALSE,
  conflevel = 95,
  p.adjust = 'fdr')
res <- res[!is.na(res$P),]
res

res <- res[order(res$LogRank, decreasing = FALSE),]
final_after <- subset(res, LogRank < 0.01)
intersect(final_after$Variable, final_wk4$Variable)
length(setdiff(final_wk4$Variable, final_base$Variable))
write.csv(final_after, "1mon expression with survival.csv")
```

## Is there any genes highly expressed between the two timepoints and indicative of survival, patients with just baseline and week4 

imported DESeq2 results of the fold change. -not valid, need fold change for each patients. 
```{r}
sdata_base <- sdata1[sdata1$TimePoint=="Base",]
sdata_mid <- sdata1[sdata1$TimePoint=="Mid",]
sdata_after <- sdata1[sdata1$TimePoint=="After",]
patient_id <- intersect(sdata_base$PatientID, sdata_mid$PatientID)
sdata_base <- sdata_base[sdata_base$PatientID %in% patient_id,]
sdata_mid <- sdata_mid[sdata_mid$PatientID %in% patient_id,]
#generate fc
exprSet <- exprSet_backup
exprSet_base <- exprSet[,colnames(exprSet) %in% sdata_base$`Sample Name`]
exprSet_mid <- exprSet[,colnames(exprSet) %in% sdata_mid$`Sample Name`]
table(rownames(exprSet_base) == rownames(exprSet_mid))
expr_fc <- log2(exprSet_mid/exprSet_base)
#generate pdata
pdata_bw<- merge(sdata_mid, pfs2, by.x = "PatientID", by.y = "id")
```
```{r}
expr_fc1 <- t(expr_fc)
coxdata <- merge(pdata_bw, expr_fc1, by.x ="Sample Name", by.y = 0)
colnames <- make.names(colnames(coxdata))
colnames(coxdata) <- colnames
coxdata[1:10,1:20]
```

```{r}
res <- RegParallel(
  data = coxdata,
  formula = 'Surv(`PFS.month`, `progression.event`) ~ [*]',
  FUN = function(formula, data)
    coxph(formula = formula,
          data = data,
          ties = 'breslow',
          singular.ok = TRUE),
  FUNtype = 'coxph',
  variables = colnames(coxdata)[15:ncol(coxdata)],
  blocksize = 2000,
  cores = 2,
  nestedParallel = FALSE,
  conflevel = 95,
  p.adjust = 'fdr')
res <- res[!is.na(res$P),]
res

res <- res[order(res$LogRank, decreasing = FALSE),]
final_fc <- res[res$LogRank < 0.01,]
length(intersect(final_fc$Variable, final_base$Variable))
length(intersect(final_base$Variable, final_wk4$Variable))
length(setdiff(final_wk4$Variable, final_base$Variable))
```

```{r}
library(survminer)
genes <- intersect(final_fc$Variable, final_wk4$Variable)
for (i in 1:length(genes)){
  survplotdata <- coxdata[,c('PFS.month', 'progression.event',paste(genes[i]))]
  summary(survplotdata)
  
  # set Z-scale cut-offs for high and low expression
  highExpr <- as.numeric(quantile(survplotdata[,paste(genes[i])], probs = 0.75))
  lowExpr <- as.numeric(quantile(survplotdata[,paste(genes[i])], probs = 0.25))
  
  survplotdata[,paste(genes[i])] <- ifelse(survplotdata[,paste(genes[i])] >= highExpr, 'High',
                                ifelse(survplotdata[,paste(genes[i])] <= lowExpr, 'Low', 'Mid'))
  
  
  # relevel the factors to have mid as the ref level
  survplotdata[,paste(genes[i])] <- factor(survplotdata[,paste(genes[i])],
                                levels = c('Mid', 'Low', 'High'))
  
  p1 <- ggsurvplot(survfit(as.formula(paste0("Surv(PFS.month, progression.event) ~", genes[i])),
                   data = survplotdata),
            data = survplotdata,
            pval = TRUE,
            break.time.by = 5,
            ggtheme = theme_minimal())
  
  print(p1)
}

```

# Pathway analysis
```{r}
library(clusterProfiler)
library(org.Hs.eg.db)

group <- "base"
gene <- final_base$Variable
gene.df <- bitr(gene, fromType = "SYMBOL",
        toType = c("ENSEMBL", "ENTREZID"),
        OrgDb = org.Hs.eg.db)

kk <- enrichKEGG(gene         = gene.df$ENTREZID,
                 organism     = 'hsa',
                 pvalueCutoff = 1)
head(kk)
```



```{r}
ggo <- enrichGO(gene     = gene.df$ENTREZID,
                OrgDb         = org.Hs.eg.db,
                keyType       = 'ENTREZID',
                ont           = "BP",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.01,
                qvalueCutoff  = 0.05,
                readable = TRUE)

head(ggo)
if(group == "fc"){
  result_fc <- ggo@result
  result_fc <- result_fc[result_fc$pvalue < 0.01, ]
}

if(group == "base"){
  result_base <- ggo@result
  result_base <- result_base[result_base$pvalue < 0.01, ]
}

if(group == "wk4"){
  result_wk4 <- ggo@result
  result_wk4 <- result_wk4[result_wk4$pvalue < 0.01, ]
}
length(intersect(result_base$ID, result_fc$ID))

```




> longitutional survival analysis, needs a little more reading. 
